---
title: "EARL Tweets"
author: "Adnan Fiaz"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message=FALSE)
library(tidyverse)
library(purrr)
library(scales)
```

Every now and then I want to stretch my data analysis muscles and create some awesome visualisations. With two out of three EARL conferences part of R history we have a readily available topic. The data is also readily available since it's Twitter data we will be using. Twitter is an amazingly rich data source and a great starting point for a data analysis (I feel there should be a awesome-twitter-blogposts list somewhere).

I was planning on using the wonderful [rtweet](http://rtweet.info/) package by Micheal Kearney but unfortunately the Twitter API doesn't provide a full history of tweets. Instead I had to revert to a Python package (*gasp*) called [GetOldTweets](https://github.com/Jefferson-Henrique/GetOldTweets-python). I strongly recommend using the official Twitter API first before going down this path.  

### The Data

```{r}
# I have used the Exporter script with the hashtags #EARLConf2017, #EARLConf and #EARL2017
tweets_df <- purrr::map_df(list.files('data', full.names = TRUE), 
                           ~ readr::read_delim(.x, delim=";", quote=""))
head(tweets_df %>% select(username, date, retweets))
```

First things first, let's get a timeline up:

```{r, echo=FALSE}

tweets_df %>% 
  mutate(day = lubridate::date(date)) %>% 
  count(day) %>% 
  ggplot(aes(x=day, y=n)) +
  geom_line() +
  labs(x='Date', y='Number of tweets', title='Number of EARL-related tweets by day') +
  scale_x_date(date_breaks="2 months", labels=date_format('%b-%y')) +
  theme_classic()
```

The hashtags I used to search tweets were generic so the results include tweets from last year's conferences. Let's zoom in on this year's conferences: EARL San Fransisco (5-7 June) and EARL London (12-14 September). They clearly explain the large peaks in the above graph.

```{r, echo=FALSE}

tweets_df %>% 
  filter(date > "2017-05-01") %>% 
  mutate(day = lubridate::date(date)) %>% 
  count(day) %>% 
  ggplot(aes(x=day, y=n)) +
  geom_line() +
  annotate("rect",xmin=lubridate::date("2017-06-05"), 
                xmax=lubridate::date("2017-06-08"), 
                ymin=0, ymax=Inf, alpha=0.2, fill="orange") +
  annotate("rect",xmin=lubridate::date("2017-09-12"), 
                xmax=lubridate::date("2017-09-15"), 
                ymin=0, ymax=Inf, alpha=0.2, fill="black") +
  labs(x='Date', y='Number of tweets', title='Number of EARL-related tweets by day') +
  scale_x_date(date_breaks="2 months", labels=date_format('%b-%y')) +
  theme_classic()

```


